Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types

### 2016-02-29 manuscript.Rnw
Typo fixes; added sitch2003

One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. One possible explanation is that gymnosperms compensate for their high leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, in order to maintain sufficient total leaf area. If this is the case, the total leaf area of plants at a standardised size should be less variable between angiosperms and gymnosperms than total leaf mass. The idea that high total leaf mass may compensate for high LMA would also explain the difference between evergreen (low LMA, low leaf mass) and deciduous (high LMA, high leaf mass) species \citep{poorter2015}. Some oft-cited studies make a different assumption: that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}. Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to to LMA.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages). In this paper we restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass; but usually for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. Our analysis has several parts; to maximise the amount of available data for each part we we used different subsets of the data, as not all variables were measured in each study. Sample sizes by PFT are summarised in Table \ref{table:counttable}.

@@

All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, re and `remake` \citep{fitzjohn2015}. In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).

@@

We found large differences between the three major PFTs in terms of above-ground biomass distribution,a s quantified by the ratio $\mfms$ (Fig. 3a). These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height for each of the PFTs. These three size-corrected $\mfms$ values were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3c), because the differences in $\mfas$ were exactly compensated by differences in LMA (see also Fig. S2 for $\afms$ as a function of plant height).

@@

We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass. \citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI). Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.  Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}. The ultimate reason LMA might effect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}. For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan. Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}. We argue that there are in fact two effects occurring simultaneously: differences in LMA cause differences in the $\mfms$ of individual plants, while differences in LL (and thereby LMA) cause differences in the LAI of whole stands.

@@

Global vegetation models assume or predict differences in biomass distribution between PFTs. However, these differences are currently poorly constrained, due to limited available empirical data. Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Indeed, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data. Our results also suggest a rethink in the way LMA is handled in most growth models. In most existing models, biomass allocation is determined first (by some means or another), and then the amount of leaf area is calculated by multiplying leaf biomass allocation with LMA. Our results suggest the chain of causation should be reversed: first decide the amount of leaf area, then let LMA determine the cost of building this leaf. Such an approach has already been implemented in at least one model of plant growth \citep{falster2016}. Other algorithms are of course possible, but this is at least one way to ensure model predictions recover the patterns reported in this paper.

>>>>

One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. One possible explanation is that gymnosperms compensate for their high leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, in order to maintain sufficient total leaf area. If this is the case, the total leaf area of plants at a standardised size should be less variable between angiosperms and gymnosperms than total leaf mass. The idea that high total leaf mass may compensate for high LMA would also explain the difference between evergreen (low LMA, low leaf mass) and deciduous (high LMA, high leaf mass) species \citep{poorter2015}. Some oft-cited studies make a different assumption: that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. Although recent large compilations on biomass distribution included data on leaf mass, they did not include measurements of leaf area so were unable to evaluate these alternatives \citep{poorter2012,poorter2015,reich2014}. Here we set out to directly test the hypothesis that across PFTs, the ratio of leaf mass to stem mass -- $\mfms$ -- is related to LMA, but the ratio of leaf area to stem mass is invariant with respect to LMA.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{falster2015}, which in total includes records for \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. The database has very limited overlap (n = 261 individuals, 1.7 \%) with the recent large compilation of \citet{poorter2015} and differs in that measurements are all for individual plants (where \citet{poorter2015} included many stand-based averages). In this paper we restrict our analysis to records that include leaf mass ($\mf$), leaf area ($\af$), above-ground woody biomass ($\ms$), plant height ($H$), and stem area measured at ground level ($\as$), or at breast height (typically 1.3m) ($\asb$) (n=\Sexpr{nrow(dataset)}). The database contains many more variables, for example root biomass; but usually for a much smaller subset of studies. Here we limit the analysis to patterns in aboveground biomass distribution. Our analysis has several parts; to maximise the amount of available data for each part we used different subsets of the data, as not all variables were measured in each study. Sample sizes by PFT are summarised in Table \ref{table:counttable}.

@@

All analyses were conducted in R 3.2.3 \citep{R-2015}, using the packages `mgcv` \citep{wood2006}, `hier.part` \citep{walsh2013}, `MuMIn` \citep{barton2015}, and `remake` \citep{fitzjohn2015}. In addition, the code replicating this analysis (and all figures) is available at \url{http://github.com/RemkoDuursma/baadanalysis} (doi: \emph{to be added}).

@@

We found large differences between the three major PFTs in terms of above-ground biomass distribution,a s quantified by the ratio $\mfms$ (Fig. 3a). These differences were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height for each of the PFTs. These three size-corrected $\mfms$ values were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit woody above-ground biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3c), because the differences in $\mfms$ were exactly compensated by differences in LMA (see also Fig. S2 for $\afms$ as a function of plant height).

@@

We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass. \citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI). Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.  Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}. The ultimate reason LMA might affect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}. For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan. Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}. We argue that there are in fact two effects occurring simultaneously: differences in LMA cause differences in the $\mfms$ of individual plants, while differences in LL (and thereby LMA) cause differences in the LAI of whole stands.

@@

Global vegetation models assume or predict differences in biomass distribution between PFTs. However, these differences are currently poorly constrained, due to limited available empirical data. Moreover, there is little consensus on how biomass distribution (and allocation) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass distribution instead of using constant allocation fractions. Indeed, the growing availability of large datasets on stand biomass and individual plant construction  \citep{falster2015,poorter2015,reich2014} suggest the time is ripe for rigorous benchmarking \citep{abramowitz2012, dekauwe2014} of GVMs against empirical data. Our results also suggest a rethink in the way LMA is handled in most growth models. In most existing models, biomass allocation is determined first (by some means or another), and then the amount of leaf area is calculated by multiplying leaf biomass allocation with LMA. Our results suggest the chain of causation should be reversed: first decide the amount of leaf area, then let LMA determine the cost of building this leaf. Such an approach has already been implemented in at least two vegetation models \citep{falster2016,sitch2003}. Other algorithms are of course possible, but this is at least one way to ensure model predictions recover the patterns reported in this paper.
